# Data

## Description
(ACS) 5-Year Housing Estimates. The data is collected and maintained by the U.S. 
Census Bureau, which conducts comprehensive surveys to assess housing 
characteristics across counties in the United States and the specific dataset we 
are using is for the years 2016-2020. The dataset is provided in a CSV format and 
includes over 240 columns with variables such as total housing units, occupancy 
rates, and housing costs. These variables are measured and updated annually, 
with detailed geographic identifiers for each county. The data is 
well-documented, and the source is reliable for understanding housing trends at 
a granular level. To import the data, I used standard data manipulation tools. 
Some challenges include the large number of columns and potential for missing 
values, which require preprocessing for effective analysis. Additionally, the 
original column names in the dataset are not clear and understandable so we need 
to rename the columns for clarity. The dataset can be accessed directly from the 
U.S. Census Bureau's data repository.

## Missing value analysis
```{r}
library(tidyverse)  
library(naniar) 
file_path <- "ACS_5YR_Housing_Estimate_Data_by_County_-8702403216850992228.csv"
housing_data <- read.csv(file_path)

# Rename specified columns explicitly
filtered_data <- housing_data %>%
  rename(
    Occupancy_Status = B25002EST1,
    Tenure_By_Household_Size = B25009EST2,
    Median_Number_Of_Rooms_By_Tenure = B25021EST1,
    Units_In_Structure = B25024EST2,
    Tenure_By_Units_In_Structure = B25032EST2,
    Tenure_By_Year_Structure_Built = B25036_OWN_80TO99,
    Median_Year_Structure_Built_By_Tenure = B25037EST3,
    Bedrooms = B25041EST2,
    Tenure_By_Bedrooms = B25042EST3,
    Contract_Rent = B25056_500TO599,
    Median_Contract_Rent = B25058EST1,
    Bedrooms_By_Gross_Rent = B25068EST2,
    Mortgage_Status_By_Median_Value = B25097EST1,
    Tenure_By_Selected_Physical_And_Financial_Conditions = B25123_TOT
  )

# Keep only the renamed columns and the original identifiers
columns_to_keep <- c(
  "Occupancy_Status",
  "Tenure_By_Household_Size",
  "Median_Number_Of_Rooms_By_Tenure",
  "Units_In_Structure",
  "Tenure_By_Units_In_Structure",
  "Tenure_By_Year_Structure_Built",
  "Median_Year_Structure_Built_By_Tenure",
  "Bedrooms",
  "Tenure_By_Bedrooms",
  "Contract_Rent",
  "Median_Contract_Rent",
  "Bedrooms_By_Gross_Rent",
  "Mortgage_Status_By_Median_Value",
  "Tenure_By_Selected_Physical_And_Financial_Conditions",
  "OBJECTID_1", "GEOID", "STATE", "STUSAB", "STATE_NAME", "COUNTY", "NAME"
)

filtered_data <- filtered_data %>%
  select(all_of(columns_to_keep))

# Print renamed and filtered columns for verification
print("Renamed and Filtered Columns:")
print(colnames(filtered_data))
```

```{r}
missing_summary <- filtered_data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Column", values_to = "MissingCount") %>%
  arrange(desc(MissingCount))

# Print missing value summary
print("Missing Values Summary:")
print(missing_summary)

# Calculate Missing Value Percentages
missing_summary <- missing_summary %>%
  mutate(MissingPercentage = (MissingCount / nrow(filtered_data)) * 100)

# Print missing percentages
print("Missing Values Percentage:")
print(missing_summary)

# Visualize Missing Values
# Heatmap of Missingness
row_missing_summary <- filtered_data %>%
  mutate(MissingCount = rowSums(is.na(.))) %>%
  group_by(MissingCount) %>%
  summarize(Count = n()) %>%
  arrange(MissingCount)

# Plot the Improved Distribution
ggplot(row_missing_summary, aes(x = MissingCount, y = Count)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black", width = 0.8) +
  geom_text(aes(label = Count), vjust = -0.3, size = 4) +
  scale_x_continuous(breaks = 0:max(row_missing_summary$MissingCount)) +
  labs(
    title = "Distribution of Missing Data Across Rows",
    subtitle = "Number of rows with varying levels of missing data",
    x = "Number of Missing Values in a Row",
    y = "Frequency (Number of Rows)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  )

ggplot(missing_summary, aes(x = reorder(Column, -MissingPercentage), y = MissingPercentage)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black") +
  geom_text(aes(label = paste0(round(MissingPercentage, 1), "%")), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(
    title = "Percentage of Missing Values by Column",
    subtitle = "Highlighting the extent of missing data in each variable",
    x = "Columns",
    y = "Percentage Missing"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  )
```