# Results
```{r}
library(tidyverse) 
library(naniar) 
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
input <- "data.csv"
data <- read.csv(input)
```

1.
```{r}
data$Median_value_for_units_with_a_mortgage <- 
  as.numeric(data$Median_value_for_units_with_a_mortgage)

state_averages <- data |>
  group_by(STATE_NAME) |>
  summarise(Median_value_for_units_with_a_mortgage = 
              mean(Median_value_for_units_with_a_mortgage, na.rm = TRUE))

top_10_states <- state_averages |>
  arrange(desc(Median_value_for_units_with_a_mortgage)) |>
  slice_head(n = 10)

bottom_10_states <- state_averages |>
  arrange(Median_value_for_units_with_a_mortgage) |>
  slice_head(n = 10)

combined_states <- bind_rows(top_10_states, bottom_10_states)

combined_states <- combined_states |>
  arrange(desc(Median_value_for_units_with_a_mortgage)) |>
  mutate(STATE_NAME = factor(STATE_NAME, levels = STATE_NAME))

ggplot(combined_states, aes(
  x = STATE_NAME, 
  y = Median_value_for_units_with_a_mortgage, 
  fill = Median_value_for_units_with_a_mortgage
)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(
    title = "Top 10/Bottom 10 States: Median Housing Values for Units With Mortgages",
    x = "State/District",
    y = "Median Housing Value"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )
```
This graph is showing the median housing trends for states and DC/Puerto Rico 
across the country. The values/trends you can see here are distributed unevenly.
This is because the coastal states or states with bigger urban areas have 
higher prices and this can be seen because of higher demand and economic status
compared to the rural states have lower values. 

2. 
```{r}
state_occupancy_data <- data |>
  mutate(Region = case_when(
    STATE_NAME %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", 
                      "Rhode Island", "Vermont", "New Jersey", "New York", 
                      "Pennsylvania", "District of Columbia") ~ "Northeast",
    STATE_NAME %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", 
                      "Minnesota", "Missouri", "Nebraska", "North Dakota", 
                      "Ohio", "South Dakota", "Wisconsin") ~ "Midwest",
    STATE_NAME %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", 
                      "Kentucky", "Louisiana", "Maryland", "Mississippi", 
                      "North Carolina", "Oklahoma", "South Carolina", 
                      "Tennessee","Texas", "Virginia", 
                      "West Virginia") ~ "South",
    STATE_NAME %in% c("Alaska", "Arizona", "California", "Colorado", "Hawaii", 
                      "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", 
                      "Utah", "Washington", "Wyoming") ~ "West",
    STATE_NAME %in% c("Puerto Rico") ~ "Puerto Rico"
  ))

region_medians <- state_occupancy_data |>
  group_by(Region) |>
  summarise(Median_value = mean(Median_value_for_units_with_a_mortgage, 
                                na.rm = TRUE)) |>
  arrange(desc(Median_value))

state_occupancy_data <- state_occupancy_data |>
  mutate(Region = factor(Region, levels = region_medians$Region))

ggplot(state_occupancy_data, aes(
  x = Region,
  y = Median_value_for_units_with_a_mortgage,
  fill = Region
)) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Distribution of Median Housing Values by Region",
    x = "Region",
    y = "Median Housing Value"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )
```
This graph shows the distribution of median housing values per region.Going off 
of the last graph we wanted to see if it was a regional trend. It can be seen 
that the northeast and the west are higher than the south and midwest. This goes
with the same idea mentioned before that economically higher places that have 
urban demand are more expensive regions compared to more rural regions. As a
note, we decided to include Washington D.C. in the northeast and include Puerto
Rico by itself.

3. 
```{r}
state_occupancy_data <- data |>
  group_by(STATE_NAME) |>
  summarise(
    Total = sum(`Occupancy_Status.Total_Housing_units.`, na.rm = TRUE),
    Occupied = sum(`Occupancy_Status.Occupied_housing_units.`, na.rm = TRUE),
    Vacant = sum(`Occupancy_Status.Vacant.`, na.rm = TRUE)
  ) |>
  pivot_longer(cols = c("Total", "Occupied", "Vacant"), 
               names_to = "Status", 
               values_to = "Units")

state_occupancy_data <- state_occupancy_data |>
  mutate(Region = case_when(
    STATE_NAME %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", 
                      "Rhode Island", "Vermont", "New Jersey", "New York", 
                      "Pennsylvania", "District of Columbia") ~ "Northeast",
    STATE_NAME %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", 
                      "Minnesota", "Missouri", "Nebraska", "North Dakota", 
                      "Ohio", "South Dakota", "Wisconsin") ~ "Midwest",
    STATE_NAME %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", 
                      "Kentucky", "Louisiana", "Maryland", "Mississippi", 
                      "North Carolina", "Oklahoma", "South Carolina", 
                      "Tennessee", "Texas", "Virginia", 
                      "West Virginia") ~ "South",
    STATE_NAME %in% c("Alaska", "Arizona", "California", "Colorado", "Hawaii", 
                      "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", 
                      "Utah", "Washington", "Wyoming") ~ "West",
    STATE_NAME %in% c("Puerto Rico") ~ "Puerto Rico"
  ))

# Filter out the "Total" category
state_occupancy_data <- state_occupancy_data |>
  filter(Status != "Total")

region_totals <- state_occupancy_data |>
  group_by(Region) |>
  summarise(Total_Units = sum(Units, na.rm = TRUE)) |>
  arrange(desc(Total_Units))

state_occupancy_data <- state_occupancy_data |>
  mutate(Region = factor(Region, levels = region_totals$Region))

ggplot(state_occupancy_data, aes(x = Region, 
                                 y = Units, fill = Status)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of Housing Units by Region and Occupancy Status",
    x = "Region",
    y = "Number of Units",
    fill = "Occupancy Status"
  ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold"))
```
This graph is explaining how different regions and their housing occupancy. We 
chose to include Puerto Rico in its own group and Washington D.C. in the 
Northeast like we did previously. The south has the highest total than the 
others by a significant amount. Surprisingly the Northeast has the least amount 
of units not including Puerto Rico by itself. This may be due to the overall 
area of the northeast being smaller than the other regions, but it is still 
surprising. The vacant is higher in rural and economically smaller areas while 
more dense places that economically higher and higher demand have less vacant.

4. 
```{r}
year_built_data <- data |>
  summarise(
    `Owner 2000+` = sum(`Owner_Tenure_By_Year_Structure_Built.Owner_occupied_housing_units_Built_2000_or_later.`, na.rm = TRUE),
    `Owner 1980-1999` = sum(`Owner_Tenure_By_Year_Structure_Built.Owner_occupied_housing_units_Built_1980_to_1999.`, na.rm = TRUE),
    `Owner 1950-1979` = sum(`Owner_Tenure_By_Year_Structure_Built.Owner_occupied_housing_units_Built_1950_to_1979.`, na.rm = TRUE),
    `Owner <1949` = sum(`Owner_Tenure_By_Year_Structure_Built.Owner_occupied_housing_units_Built_1949_or_earlier.`, na.rm = TRUE)
  ) |>
  pivot_longer(cols = everything(), names_to = "Category", values_to = "Units") |>
  separate(Category, into = c("Tenure", "Period"), sep = " ")

ggplot(year_built_data, aes(
  x = Period, 
  y = Units, 
  group = 1
)) +
  geom_line(size = 1.2, color = "orchid2") +
  geom_point(size = 3, color = "orchid2") +
  geom_text(aes(label = scales::comma(Units)), 
            vjust = -0.5, size = 3, show.legend = FALSE) +
  labs(
    title = "Housing Units Built Over Time",
    subtitle = "Trends in owner-occupied housing units across periods",
    x = "Period",
    y = "Number of Units"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 14, hjust = 0.5, face ="bold"),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.position = "none"
  )
```
In this graph the trend of when units were built over the countries history. 
We wanted to observe this because we had seen all the reports about home 
construction being down. From the graph we can now see that the number of 
new houses being built has been consistently dropping since the baby boom. 
This could be due to economic trends and how the country is growing.


5. 
```{r}
filtered_data <- data |>
  filter(Median_Year_Structure_Built_By_Tenure.Median_Year_Structure_Built_for_Renter_Occupied_Units. >= 1990 & 
         Median_Year_Structure_Built_By_Tenure.Median_Year_Structure_Built_for_Renter_Occupied_Units. <= 2022)

ggplot(filtered_data, 
       aes(x = reorder(STATE_NAME, Median_Year_Structure_Built_By_Tenure.Median_Year_Structure_Built_for_Renter_Occupied_Units., 
                       median),
           y = Median_Year_Structure_Built_By_Tenure.Median_Year_Structure_Built_for_Renter_Occupied_Units.)) +
  geom_boxplot(outlier.alpha = 0.3, fill = "skyblue", alpha = 0.6) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Variance in Year Built for Renter-Occupied Units by State (1990+)",
    x = "State (Ordered by Median Year Built)",
    y = "Year Built"
  ) +
  theme(
    plot.title = element_text(size = 11, hjust = 0.5, face = "bold"),
    axis.text.y = element_text(size = 8, hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.margin = margin(20, 50, 20, 50)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +
  coord_flip()
```
Continuing on the findings from the previous graph where we looked at 
owner-occupied housing building trends we wanted to now look at the 
renter-occupied build dates but by state. We can see here that states in the 
south and west have newer houses due to growing population and development and 
midwest/northeast houses are on the older side. We chose to show states whose 
median year built was beyond 1990 and the graph shows how only a few states even
have a median year built beyond 1990. This once again, reinforces the idea that
house construction has been continously declining.

6. 

```{r}
occupancy_data <- data.frame(
  Category = c("Occupied Housing Units", "Vacant Housing Units"),
  Values = c(
    sum(data$`Occupancy_Status.Occupied_housing_units.`, na.rm = TRUE),
    sum(data$`Occupancy_Status.Vacant.`, na.rm = TRUE)
  )
) |>
  mutate(Percentage = round(Values / sum(Values) * 100, 1)) 

ggplot(occupancy_data, aes(x = "", y = Values, fill = Category)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(Percentage, "%")), 
            position = position_stack(vjust = 0.5), 
            size = 5, color = "white") + # Add percentage labels
  theme_minimal(base_size = 16) +
  labs(title = "Occupied vs Vacant Housing Units", fill
       = "Housing Type") +
  theme(axis.text = element_blank(), 
        axis.title = element_blank(), 
        panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```
This graph was created to go with the housing trend to see overall how the 
distribution to occupied to vacant is. We can see here that the majority is 
obviously occupied and this shoes the demand for housing across the country is
strong but the vacancies might be in specific areas. 

7. 

```{r}
structure_data <- data %>%
  summarise(
    Detached = sum(`Units_In_Structure.1unit_detached_structure.`, na.rm = TRUE),
    Attached = sum(`Units_In_Structure.1unit_attached_structure.`, na.rm = TRUE),
    `2-4 Units` = sum(`Units_In_Structure.2_to_4_units.`, na.rm = TRUE),
    `5-19 Units` = sum(`Units_In_Structure.5_to_19_units.`, na.rm = TRUE),
    `20+ Units` = sum(`Units_In_Structure.20_or_more_units.`, na.rm = TRUE),
    Mobile = sum(`Units_In_Structure.Mobile_home_Boat_RV_van_et.`, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(), names_to = "Structure_Type", values_to = "Units")

structure_data <- structure_data %>%
  mutate(Structure_Type = forcats::fct_reorder(Structure_Type, Units, .desc = TRUE))

# Plot
ggplot(structure_data, aes(x = Structure_Type, y = Units, fill = Structure_Type)) +
  geom_bar(stat = "identity") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Housing Units by Structure Type (Ordered by Frequency)",
    x = "Structure Type",
    y = "Number of Units"
  ) +
  theme(legend.position = "none")
```
This graph we wanted to see what kind of structure was being built as we explore
more about housing types. We can see that detached by far is dominated the
market. This shows that people and families are preferring standalone homes in 
all types of regions. From what we got gathered earlier though, most of these
detached houses are very old houses.

8. 
```{r}
bedroom_data <- data |>
  summarise(
    `Owner No Bedroom` = 
      sum(`Tenure_By_Bedrooms.Owner_occupied_housing_units_with_No_bedroom.`, 
          na.rm = TRUE),
    `Owner 1 Bedroom` = 
      sum(`Tenure_By_Bedrooms.Owner_occupied_housing_units_with_1_bedroom.`, 
          na.rm = TRUE),
    `Owner 2 Bedrooms` = 
      sum(`Tenure_By_Bedrooms.Owner_occupied_housing_units_with_2_bedrooms.`, 
          na.rm = TRUE),
    `Owner 3+ Bedrooms` = 
      sum(`Tenure_By_Bedrooms.Owner_occupied_housing_units_with_3_or_more_bedrooms.`, 
          na.rm = TRUE),
    `Renter No Bedroom` = 
      sum(`Tenure_By_Bedrooms.Renter_occupied_housing_units_with_No_bedroom.`, 
          na.rm = TRUE),
    `Renter 1 Bedroom` = 
      sum(`Tenure_By_Bedrooms.Renter_occupied_housing_units_with_1_bedroom.`, 
          na.rm = TRUE),
    `Renter 2 Bedrooms` = 
      sum(`Tenure_By_Bedrooms.Renter_occupied_housing_units_with_2_bedrooms.`, 
          na.rm = TRUE),
    `Renter 3+ Bedrooms` = 
      sum(`Tenure_By_Bedrooms.Renter_occupied_housing_units_with_3_or_more_bedrooms.`, 
          na.rm = TRUE)
  ) |>
  pivot_longer(cols = everything(), 
               names_to = "Category", 
               values_to = "Units") |>
  separate(Category, into = c("Tenure", "Bedrooms"), sep = " ")

ggplot(bedroom_data, aes(x = Bedrooms, y = Units, fill = Tenure)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal(base_size = 14) +
  labs(title = "Bedrooms Distribution by Tenure", x = "Number of Bedrooms", 
       y = "Number of Units", fill = "Tenure")
```
This graph was to follow up on the last graph and wanted to see the size of the 
houses that were built and if it had to do with the type of ownership. It can
be seen here that units with 3+ bedrooms are mostly owned instead of rented. The
renters seem to prefer smaller places like 1 or 2 bedrooms for affordability and 
tenant needs. This makes sense because people in need of more bedrooms typically
tend to move into their own house rather than rent from someone else. 

9.
```{r}
# Aggregate the data to show the distribution of rent ranges
rent_distribution <- data |>
  pivot_longer(
    cols = starts_with("Contract_Rent"),
    names_to = "Rent_Range",
    values_to = "Count"
  ) |>
  mutate(
    Rent_Range = case_when(
      str_detect(Rent_Range, "500_to_599") ~ "$500-$599",
      str_detect(Rent_Range, "600_to_699") ~ "$600-$699",
      str_detect(Rent_Range, "700_to_799") ~ "$700-$799",
      str_detect(Rent_Range, "800_to_899") ~ "$800-$899",
      str_detect(Rent_Range, "900_to_999") ~ "$900-$999",
      str_detect(Rent_Range, "1000_to_1249") ~ "$1000-$1249",
      str_detect(Rent_Range, "1250_to_1499") ~ "$1250-$1499",
      str_detect(Rent_Range, "1500_to_1999") ~ "$1500-$1999",
      str_detect(Rent_Range, "2000_or_more") ~ "$2000+",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(Rent_Range)) |>
  group_by(Rent_Range) |>
  summarise(Count = sum(Count, na.rm = TRUE), .groups = "drop") |>
  mutate(Rent_Range = factor(Rent_Range, levels = c(
    "$500-$599", "$600-$699", "$700-$799", "$800-$899", "$900-$999", 
    "$1000-$1249", "$1250-$1499", "$1500-$1999", "$2000+"
  )))

# Create a bar chart showing the distribution of rent ranges
ggplot(rent_distribution, aes(x = Rent_Range, y = Count, fill = Rent_Range)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(
    title = "Distribution of Rent Prices Accross the US",
    x = "Rent Range",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  ) +
  scale_fill_brewer(palette = "Set3")
```
This graph shows the different ranges of rent since we are curious how the price
varies throughout the country which is more prevalent. We can see here that
1000-1249$ is the highest meaning the middle class is probably the most 
common renters but the rest are pretty uniform. However, we were surprised to
see that the distribution of rent prices is fairly even.


10. 
```{r}
ownership_data_by_state <- data |>
  group_by(STATE_NAME) |>
  summarise(
    Owner_Occupied = sum(`Tenure_By_Household_Size.Owner_Occupied.`, 
                         na.rm = TRUE),
    Renter_Occupied = sum(`Tenure_By_Household_Size.Renter_Occupied.`, 
                          na.rm = TRUE)
  ) |>
  pivot_longer(cols = c(Owner_Occupied, Renter_Occupied), 
               names_to = "Ownership", values_to = "Units")

ownership_diff <- ownership_data_by_state |>
  pivot_wider(names_from = Ownership, values_from = Units) |>
  mutate(Difference = abs(Owner_Occupied - Renter_Occupied))

top_states <- ownership_diff |>
  arrange(Difference) |>
  slice_head(n = 10)


ggplot(top_states, aes(x = reorder(STATE_NAME, -Difference))) +
  geom_point(aes(y = Owner_Occupied, color = "Owner Occupied"), size = 3) +
  geom_point(aes(y = Renter_Occupied, color = "Renter Occupied"), size = 3) +
  geom_segment(aes(xend = STATE_NAME, 
                   y = Owner_Occupied, 
                   yend = Renter_Occupied), 
               color = "gray") +
  coord_flip() +
  labs(
    title = "Top 10 States with Smallest Difference in Ownership vs Renting",
    x = "State",
    y = "Proportion of Units",
    color = "Ownership Type"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_text(size = 10),
    plot.title = element_text(size = 12, face = "bold")
  )
```
This graph was created to see which 10 states within the country had the 
smallest disparity between owners and renter percentage. We ordered the ten 
states from  those with the lowest differences between owners and renters to 
those with the highest. A huge surprise from this is that in Washington D.C.
there are actually more people renting a place than owning a place.